---
title: "Upload guidelines"
output: 
  md_document:
    df_print:
      "kable"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.table.format = 'markdown')
```

## Upload guidelines

Data in the form of a linelist and contacts list can be uploaded on this page. Templates can be downloaded containing a simple example which can be run through the app. Guidelines for adding your own data to these templates are shown below.

## Linelist

The linelist details the information on each individual. This is matched to the contacts sheet by **id**. The order of the columns does not matter. **When you have filled in the new data, save as a .csv ready to upload.**

### Mandatory fields

#### id

This is the **unique** identifier of each individual. It should match exactly with their entry (if they have one) in the contacts list. For example, the identifier could have the form "EG1".

#### reported_onset_date

This is the reported date of symptom onset. It should always be entered in the form **dd/mm/yy**, mm/dd/yy is also accepted.

#### death_date

This is the reported date of death. It should always be entered in the form **dd/mm/yy**, mm/dd/yy is also accepted.

### Optional fields

#### bleeding_at_reported_onset

Whether a person was bleeding^ at onset. This should take the value **TRUE** or **FALSE**; empty cells will be interpreted as **FALSE**.

#### diarrhea_at_reported_onset

Whether a person had diarrhea at onset. This should take the value **TRUE** or **FALSE**; empty cells will be interpreted as **FALSE**.

#### Other

Any number of extra columns may be added, such as Age or Sex. If dates, they should have the form **dd/mm/yy**, mm/dd/yy is also accepted.

## Contacts

This details the links between people. One line corresponds to one connection. **from** and **to** should only take **id** that are present in the linelist and the **id** should be in exactly the same format. The order of the columns does not matter. **When you have filled in the new data, save as a .csv ready to upload.**.

### Mandatory fields

#### from
The source of infection. This should take the unique **id** of the individual.

#### to 
The individual who is infected by the source. This should take the unique **id** of the individual. The app will accept multiple sources of infection for an individual, just add another row for the new transmission link.

### Optional fields

#### contact_of_typex

Any number of extra columns may be added to describe the **type** of link, for example **in_funeral** would indicate the transmission occurred at a funeral. All additional columns must have entries **TRUE**, **FALSE** or be left empty; empty cells are interpreted as **FALSE**. A more specific example is as follows.

```{r  contact_eg, echo = FALSE, results='asis'}
library(pander)
df = data.frame("from" = c("EG1","EG1", "EG2", "EG3"), 
                           "to" = c("EG2", "EG4", "EG5", "EG6") ,
                           "in_funeral" = c("TRUE", "FALSE", "FALSE", 
                                                  "FALSE"),
                           "in_health_facility" = c("FALSE", "FALSE", "TRUE", 
                                                  "FALSE") )
pandoc.table(df)
```

In the above, transmission from **EG1** to **EG2** occurred at a funeral and transmission from **EG2** to **EG5** occurred in a health facility.